Historical aerial surveys map long-term changes of forest cover and structure in the central Congo Basin

### 2020-01-24 manuscript/main.Rmd
revision

Tropical ecosystem services are severely impacted by deforestation and forest degradation [@hansen2013; @houghton2012; @tyukavina2015]. Not only does tropical forest Land-Use and Land-Cover Change (LULCC) constitute 10 to 15% of the total global carbon emissions [@vanderwerf2009], changes in forest fragmentation affect the forest structure and function [@fauset2017; @brinck2017; @didham1999]. Strong fragmentation effects decreases the number of large trees along forest edges [@laurance2000; @magnago2015], while species composition and biodiversity are equally negatively affected [@poorter2006; @barlow2016; @vandeperre2018]. Estimates show that 31% of carbon emissions are caused by edge effects alone [@brinck2017].

Accurate estimates of LULCC and forest canopy structure are therefore imperative to estimate carbon emissions and other ecosystem services [@houghton2012; @hansen2013]. Remote sensing products have been key inputs in LULCC assessments as they provide accurate spatial information to help estimate carbon emissions [@mitchard2018; @hansen2013]. More so, high resolution aerial images provide scientists tools to monitor forest extent, structure and carbon emissions as canopy texture is linked to above ground biomass [@Ploton2012; @Couteron2005; @Barbier2010]. Yet, most of these estimates are limited in time to recent decades [@defries2002; @houghton2012; @hansen2013; @achard2014].

Historical estimates **of Land-Use and Land-Cover (LULC)**, in the pre-satellite era (< 1972), exist but generally rely on non-spatially explicit data (i.e socio-economic data) [@houghton2012; @ramankutty1999; @defries2002; @sader1988]. Efforts have been made to use other geo-spatial data sources such as historical maps [@willcock2016], declassified CORONA satellite surveillance data across the US and central Brazil [@song2015] as well as aerial surveys in post World War II German [@nita2018]. Survey data across the African continent is less common, inaccessible or both. Some studies do exist, as @buitenwerf2012 and @hudak1998 used aerial survey images to map vegetation changes in South African savannas, whilst @frankl2015 and @nyssen2016 mapped the Ethiopian highlands of 1930's.

Across the central Congo Basin most of these historical images were collected within the context of national cartographic efforts by the "Institut Géographique du Congo Belge" in Kinshasa (then Léopoldville), DR Congo. Despite the existence of large archives of aerial survey imagery (Figure 1, Appendix Figure 3), as of yet, no studies have valorized these data. The lack of a consistent valorization effort is unfortunate as the African rainforest is the second largest on Earth and covers ~630 million ha, representing up to 66 Pg of carbon storage [@Lewis2009b], and currently loses forest at an increasing pace [@butsic2015]. Given the impact of LULCC on the structure and functioning of central African tropical forests, and their influence on both carbon dynamics [@Kearsley2013] and biodiversity [@vandeperre2018], accurate long term reporting of historical forest cover warrants more attention [@willcock2016].

Here, we use a combination of historical aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin, effectively linking the start of the antrophocene [@lewis2015] with current assessments. Yangambi was, and remains, a focal center of forest and agricultural research and development in the central Congo Basin. Past research in the region allows for thorough assessment of land-use and land-cover change from a multi-disciplinary point of view, confronting us with complex deforestation and land-use patterns.

We leverage Structure-from-Motion to generate a large orthomosaic of historical imagery and develop a Convolutional Neural Network based forest cover mapping approach, based upon a semi-supervised generated dataset extensively leveraging data augmentation. Our methodology aims to provides a historical insight into important LULCC spatial patterns in Yangambi, such as fragmentation and edge complexity. We further contextualize the influence of changes in the forest's life history on past and current research into Above Ground Carbon (AGC) storage [@Kearsley2013] and biodiversity [@vandeperre2018] in the central Congo Basin. Our fast scalable mapping approach for historical aerial survey data, using limited supervised input, would further support long-term land-use and land-cover change analysis across the central Congo Basin.

@@

We automatically delineated all natural forest in the historical data, thus excluding tree plantations, thinned or deteriorated forest stands which **showed visible canopy cover loss**, fields and buildings. We used the Unet Convolutional Neural Net (CNN, @ronneberger2015) architecture implemented in Keras [@chollet2015] with an efficientnetb3 pre-processing backbone [@yakubovskiy2019] running on TensorFlow [@martinabadi2015] to train a binary classifier (i.e. forest or non-forested). Training data were collected from the orthomosaic by randomly selecting 513 pixel square tiles from **locations within** homogeneous forested or non-forested polygons in the historical orthomosaic **(Figure 5). Separate polygons were selected for training, testing and validation purposes. Validation polygons were sampled 300 times, while both testing and validation polygons were sampled at 100 random locations. Tiles extracted from locations close to the polygon border at times contained mixed cover types. Tiles with mixed cover types were removed from the list of source tiles (Table 1).** Homogeneous **source** tiles were combined in synthetic landscapes using a random gaussian field based binary mask (Figure 3). We generated 5000 synthetic landscapes for training, while 500 landscapes were generated for both the validation and the testing datasets for a total of 6000 synthetic landscapes. In order to limit stitch line misclassifications, along the seams of mosaicked images, we created synthetic landscapes with different forest tiles to mimick forest texture transitions. We applied this technique to 10% of the generated synthetic landscapes **(across training, validation and testing data)**.

@@

To quantify changes in the structure of forest cover and its disturbances we used spatial pattern analysis landscape fragmentation metrics [@dale1999]. We report the ratio of edge to area and the fractal index to quantify landscape complexity of forest disturbances. Statistics were calculated for all forest disturbance patches larger than 1 ha and smaller than the 95^th^ percentile of the patch size distribution using the R package landscapemetrics [@hesselbarth2019]. We provide mean and standard deviation on edge, area, their ratio and fractal dimension for both the historical and contemporary @hansen2013 forest cover maps.

@@

Over the past half century there has been a clear shift in land use in Yangambi (Figure 5). Land use has shifted away from for example a regular fishbone deforestation pattern that emerges when (large scale) agricultural interests dominate the landscape [@arima2015], to a more fragmented landscape (Figure 5D). The latter former? is consistent with historical land management at the time of the aerial survey [@bustillo2018]. These regular patterns have since been reversed because of a decrease in large scale intensive agriculture, replaced by ad-hoc small scale subsistence farming with large perimeter to area relationships (i.e. ragged edges). Consequently, edge effects in the current landscape are far more pronounced than in the historical scene.

@@

Given the impact of tropical forest disturbances on atmospheric carbon emissions, biodiversity and ecosystem productivity accurate long term reporting of LULCC is an imperative. Our analysis of historical aerial survey images (1958) of the Central Congo Basin provides a window into the state of the forest at the start of the anthropocene. The use of a CNN based LULC classifier, using synthetic landscapes based image augmentation, provides a robust semi-supervised solution which scales across space and time, even for image scenes with inconsistent illumination or sharpness. Combined with contemporary remote sensing data we have shown that historical aerial survey data can be used to quantify long-term changes in LULC and AGC. We showed a shift from previously highly structured industrial deforestation of large areas **for plantation purposes**, to discrete smallholder clearing for farming, increasing landscape fragmentation but also opportunties for substantial regrowth. Efforts to quantify canopy texture features and their link to AGC had limited to no success. Our analysis provides insights into the rate at which deforestation and reforestation has taken place over a multi-decadal scale in the central Congo basin. As such, it provides a useful historical context while interpreting past and ongoing forest research in the area.

>>>>

**Tropical ecosystem services are severely impacted by deforestation and forest degradation [@hansen2013; @houghton2012; @tyukavina2015]. Not only does tropical forest Land-Use and Land-Cover Change (LULCC) constitute 10 to 15% of the total global carbon emissions [@vanderwerf2009], changes in forest fragmentation affect the forest structure and function [@fauset2017; @brinck2017; @didham1999]. Strong fragmentation effects decreases the number of large trees along forest edges [@laurance2000; @magnago2015], while species composition and biodiversity are equally negatively affected [@poorter2006; @barlow2016; @vandeperre2018]. Estimates show that 31% of carbon emissions are caused by edge effects alone [@brinck2017].**

**Accurate estimates of LULCC and forest canopy structure are therefore imperative to estimate carbon emissions and other ecosystem services [@houghton2012; @hansen2013]. Remote sensing products have been key inputs in LULCC assessments as they provide accurate spatial information to help estimate carbon emissions [@mitchard2018; @hansen2013]. More so, high resolution aerial images provide scientists tools to monitor forest extent, structure and carbon emissions as canopy texture is linked to above ground biomass [@Ploton2012; @Couteron2005; @Barbier2010]. Yet, most of these estimates are limited in time to recent decades [@defries2002; @houghton2012; @hansen2013; @achard2014].**

**Historical estimates of Land-Use and Land-Cover (LULC), in the pre-satellite era (< 1972), exist but generally rely on non-spatially explicit data (i.e socio-economic data) [@houghton2012; @ramankutty1999; @defries2002; @sader1988]. Efforts have been made to use other geo-spatial data sources such as historical maps [@willcock2016], declassified CORONA satellite surveillance data across the US and central Brazil [@song2015] as well as aerial surveys in post World War II German [@nita2018]. Survey data across the African continent is less common, inaccessible or both. Some studies do exist, as @buitenwerf2012 and @hudak1998 used aerial survey images to map vegetation changes in South African savannas, whilst @frankl2015 and @nyssen2016 mapped the Ethiopian highlands of 1930's.**

**Across the central Congo Basin most of these historical images were collected within the context of national cartographic efforts by the "Institut Géographique du Congo Belge" in Kinshasa (then Léopoldville), DR Congo. Despite the existence of large archives of aerial survey imagery (Figure 1, Appendix Figure 3), as of yet, no studies have valorized these data. The lack of a consistent valorization effort is unfortunate as the African rainforest is the second largest on Earth and covers ~630 million ha, representing up to 66 Pg of carbon storage [@Lewis2009b], and currently loses forest at an increasing pace [@butsic2015]. Given the impact of LULCC on the structure and functioning of central African tropical forests, and their influence on both carbon dynamics [@Kearsley2013] and biodiversity [@vandeperre2018], accurate long term reporting of historical forest cover warrants more attention [@willcock2016].**

**Here, we use a combination of historical aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin, effectively linking the start of the antrophocene [@lewis2015] with current assessments. Yangambi was, and remains, a focal center of forest and agricultural research and development in the central Congo Basin. Past research in the region allows for thorough assessment of land-use and land-cover change from a multi-disciplinary point of view, confronting us with complex deforestation and land-use patterns.**

**We leverage Structure-from-Motion to generate a large orthomosaic of historical imagery and develop a Convolutional Neural Network based forest cover mapping approach, based upon a semi-supervised generated dataset extensively leveraging data augmentation. Our methodology aims to provides a historical insight into important LULCC spatial patterns in Yangambi, such as fragmentation and edge complexity. We further contextualize the influence of changes in the forest's life history on past and current research into Above Ground Carbon (AGC) storage [@Kearsley2013] and biodiversity [@vandeperre2018] in the central Congo Basin. Our fast scalable mapping approach for historical aerial survey data, using limited supervised input, would further support long-term land-use and land-cover change analysis across the central Congo Basin.**

@@

We automatically delineated all natural forest in the historical data, thus excluding tree plantations, thinned or deteriorated forest stands which **showed visible canopy cover loss**, fields and buildings. We used the Unet Convolutional Neural Net (CNN, @ronneberger2015) architecture implemented in Keras [@chollet2015] with an efficientnetb3 pre-processing backbone [@yakubovskiy2019] running on TensorFlow [@martinabadi2015] to train a binary classifier (i.e. forest or non-forested). Training data were collected from the orthomosaic by randomly selecting 513 pixel square tiles from **locations within** homogeneous forested or non-forested polygons in the historical orthomosaic **(Figure 5). Separate polygons were selected for training, testing and validation purposes. Validation polygons were sampled 300 times, while both testing and validation polygons were sampled at 100 random locations. Tiles extracted from locations close to the polygon border at times contained mixed cover types. Tiles with mixed cover types were removed from the list of source tiles (Table 1).** Homogeneous **source** tiles were combined in synthetic landscapes using a random gaussian field based binary mask (Figure 3). We generated 5000 synthetic landscapes **(balancing forest and non-forest classes)** for training, while 500 landscapes were generated for both the validation and the testing datasets for a total of 6000 synthetic landscapes. In order to limit stitch line misclassifications, along the seams of mosaicked images, we created synthetic landscapes with different forest tiles to mimick forest texture transitions. We applied this technique to 10% of the generated synthetic landscapes **(across training, validation and testing data)**.

@@

To quantify changes in the structure of forest cover and its disturbances we used spatial pattern analysis landscape (fragmentation) metrics [@dale1999]. **Landscape metrics provide a mathematical framework for the analysis of discrete land‐cover classes and allows us to capture their composition and configuration. These metrics are therefore commonly used to compare how landscapes change over time [@hesselbarth2019]. In particular, fractals provide a way to quantify complex natural landscapes, including their self-similarity, across scales [@Li2000; @Mandelbrot1977]** We report the ratio of edge to area and the fractal dimension to quantify landscape complexity of forest disturbances. **A fractal dimension closer to 2 suggest a more complex (fragmented) landscape.**

Statistics were calculated for all forest disturbance patches larger than 1 ha and smaller than the 95^th^ percentile of the patch size distribution using the R package landscapemetrics [@hesselbarth2019]. We provide mean and standard deviation on edge, area, their ratio and fractal dimension for both the historical and contemporary @hansen2013 forest cover maps.

@@

Over the past half century there has been a clear shift in land use in Yangambi (Figure 5). Land use has shifted away from for example a regular fishbone deforestation pattern that emerges when (large scale) agricultural interests dominate the landscape [@arima2015], to a more fragmented landscape (Figure 5D). The former is consistent with historical land management at the time of the aerial survey [@bustillo2018]. These regular patterns have since been reversed because of a decrease in large scale intensive agriculture, replaced by ad-hoc small scale subsistence farming with large perimeter to area relationships (i.e. ragged edges). Consequently, edge effects in the current landscape are far more pronounced than in the historical scene.

@@

**Given the impact of tropical forest disturbances on atmospheric carbon emissions, biodiversity and ecosystem productivity accurate long term reporting of LULCC is an imperative. Our analysis of historical aerial survey images (1958) of the Central Congo Basin provides a window into the state of the forest at the start of the anthropocene. The use of a CNN based LULC classifier, using synthetic landscapes based image augmentation, provides a robust semi-supervised solution which scales across space and time, even for image scenes with inconsistent illumination or sharpness. Combined with contemporary remote sensing data we have shown that historical aerial survey data can be used to quantify long-term changes in LULC and AGC. We showed a shift from previously highly structured industrial deforestation of large areas for plantation purposes, to discrete smallholder clearing for farming, increasing landscape fragmentation but also opportunties for substantial regrowth. Efforts to quantify canopy texture features and their link to AGC had limited to no success. Our analysis provides insights into the rate at which deforestation and reforestation has taken place over a multi-decadal scale in the central Congo basin. As such, it provides a useful historical context while interpreting past and ongoing forest research in the area.**

### 2020-01-24 manuscript/main.tex
revision

Tropical ecosystem services are severely impacted by deforestation and
forest degradation \citep{hansen2013, houghton2012, tyukavina2015}. Not
only does tropical forest Land-Use and Land-Cover Change (LULCC)
constitute 10 to 15\% of the total global carbon emissions
\citep{vanderwerf2009}, changes in forest fragmentation affect the
forest structure and function
\citep{fauset2017, brinck2017, didham1999}. Strong fragmentation effects
decreases the number of large trees along forest edges
\citep{laurance2000, magnago2015}, while species composition and
biodiversity are equally negatively affected
\citep{poorter2006, barlow2016, vandeperre2018}. Estimates show that
31\% of carbon emissions are caused by edge effects alone
\citep{brinck2017}.

Accurate estimates of LULCC and forest canopy structure are therefore
imperative to estimate carbon emissions and other ecosystem services
\citep{houghton2012, hansen2013}. Remote sensing products have been key
inputs in LULCC assessments as they provide accurate spatial information
to help estimate carbon emissions \citep{mitchard2018, hansen2013}. More
so, high resolution aerial images provide scientists tools to monitor
forest extent, structure and carbon emissions as canopy texture is
linked to above ground biomass
\citep{Ploton2012, Couteron2005, Barbier2010}. Yet, most of these
estimates are limited in time to recent decades
\citep{defries2002, houghton2012, hansen2013, achard2014}.

Historical estimates \textbf{of Land-Use and Land-Cover (LULC)}, in the
pre-satellite era (\textless{} 1972), exist but generally rely on
non-spatially explicit data (i.e socio-economic data)
\citep{houghton2012, ramankutty1999, defries2002, sader1988}. Efforts
have been made to use other geo-spatial data sources such as historical
maps \citep{willcock2016}, declassified CORONA satellite surveillance
data across the US and central Brazil \citep{song2015} as well as aerial
surveys in post World War II German \citep{nita2018}. Survey data across
the African continent is less common, inaccessible or both. Some studies
do exist, as \citet{buitenwerf2012} and \citet{hudak1998} used aerial
survey images to map vegetation changes in South African savannas,
whilst \citet{frankl2015} and \citet{nyssen2016} mapped the Ethiopian
highlands of 1930's.

Across the central Congo Basin most of these historical images were
collected within the context of national cartographic efforts by the
``Institut Géographique du Congo Belge'' in Kinshasa (then
Léopoldville), DR Congo. Despite the existence of large archives of
aerial survey imagery (Figure 1, Appendix Figure 3), as of yet, no
studies have valorized these data. The lack of a consistent valorization
effort is unfortunate as the African rainforest is the second largest on
Earth and covers \textasciitilde{}630 million ha, representing up to 66
Pg of carbon storage \citep{Lewis2009b}, and currently loses forest at
an increasing pace \citep{butsic2015}. Given the impact of LULCC on the
structure and functioning of central African tropical forests, and their
influence on both carbon dynamics \citep{Kearsley2013} and biodiversity
\citep{vandeperre2018}, accurate long term reporting of historical
forest cover warrants more attention \citep{willcock2016}.

Here, we use a combination of historical aerial photography and
contemporary remote sensing data to map long-term changes in the extent
and structure of the tropical forest surrounding Yangambi (DR Congo), in
the central Congo Basin, effectively linking the start of the
antrophocene \citep{lewis2015} with current assessments. Yangambi was,
and remains, a focal center of forest and agricultural research and
development in the central Congo Basin. Past research in the region
allows for thorough assessment of land-use and land-cover change from a
multi-disciplinary point of view, confronting us with complex
deforestation and land-use patterns.

We leverage Structure-from-Motion to generate a large orthomosaic of
historical imagery and develop a Convolutional Neural Network based
forest cover mapping approach, based upon a semi-supervised generated
dataset extensively leveraging data augmentation. Our methodology aims
to provides a historical insight into important LULCC spatial patterns
in Yangambi, such as fragmentation and edge complexity. We further
contextualize the influence of changes in the forest's life history on
past and current research into Above Ground Carbon (AGC) storage
\citep{Kearsley2013} and biodiversity \citep{vandeperre2018} in the
central Congo Basin. Our fast scalable mapping approach for historical
aerial survey data, using limited supervised input, would further
support long-term land-use and land-cover change analysis across the
central Congo Basin.

@@

We automatically delineated all natural forest in the historical data,
thus excluding tree plantations, thinned or deteriorated forest stands
which \textbf{showed visible canopy cover loss}, fields and buildings.
We used the Unet Convolutional Neural Net (CNN, \citet{ronneberger2015})
architecture implemented in Keras \citep{chollet2015} with an
efficientnetb3 pre-processing backbone \citep{yakubovskiy2019} running
on TensorFlow \citep{martinabadi2015} to train a binary classifier
(i.e.~forest or non-forested). Training data were collected from the
orthomosaic by randomly selecting 513 pixel square tiles from
\textbf{locations within} homogeneous forested or non-forested polygons
in the historical orthomosaic \textbf{(Figure 5). Separate polygons were
selected for training, testing and validation purposes. Validation
polygons were sampled 300 times, while both testing and validation
polygons were sampled at 100 random locations. Tiles extracted from
locations close to the polygon border at times contained mixed cover
types. Tiles with mixed cover types were removed from the list of source
tiles (Table 1).} Homogeneous \textbf{source} tiles were combined in
synthetic landscapes using a random gaussian field based binary mask
(Figure 3). We generated 5000 synthetic landscapes for training, while
500 landscapes were generated for both the validation and the testing
datasets for a total of 6000 synthetic landscapes. In order to limit
stitch line misclassifications, along the seams of mosaicked images, we
created synthetic landscapes with different forest tiles to mimick
forest texture transitions. We applied this technique to 10\% of the
generated synthetic landscapes \textbf{(across training, validation and
testing data)}.

@@

To quantify changes in the structure of forest cover and its
disturbances we used spatial pattern analysis landscape fragmentation
metrics \citep{dale1999}. We report the ratio of edge to area and the
fractal index to quantify landscape complexity of forest disturbances.
Statistics were calculated for all forest disturbance patches larger
than 1 ha and smaller than the 95\textsuperscript{th} percentile of the
patch size distribution using the R package landscapemetrics
\citep{hesselbarth2019}. We provide mean and standard deviation on edge,
area, their ratio and fractal dimension for both the historical and
contemporary \citet{hansen2013} forest cover maps.

@@

Over the past half century there has been a clear shift in land use in
Yangambi (Figure 5). Land use has shifted away from for example a
regular fishbone deforestation pattern that emerges when (large scale)
agricultural interests dominate the landscape \citep{arima2015}, to a
more fragmented landscape (Figure 5D). The latter former? is consistent
with historical land management at the time of the aerial survey
\citep{bustillo2018}. These regular patterns have since been reversed
because of a decrease in large scale intensive agriculture, replaced by
ad-hoc small scale subsistence farming with large perimeter to area
relationships (i.e.~ragged edges). Consequently, edge effects in the
current landscape are far more pronounced than in the historical scene.

@@

Given the impact of tropical forest disturbances on atmospheric carbon
emissions, biodiversity and ecosystem productivity accurate long term
reporting of LULCC is an imperative. Our analysis of historical aerial
survey images (1958) of the Central Congo Basin provides a window into
the state of the forest at the start of the anthropocene. The use of a
CNN based LULC classifier, using synthetic landscapes based image
augmentation, provides a robust semi-supervised solution which scales
across space and time, even for image scenes with inconsistent
illumination or sharpness. Combined with contemporary remote sensing
data we have shown that historical aerial survey data can be used to
quantify long-term changes in LULC and AGC. We showed a shift from
previously highly structured industrial deforestation of large areas
\textbf{for plantation purposes}, to discrete smallholder clearing for
farming, increasing landscape fragmentation but also opportunties for
substantial regrowth. Efforts to quantify canopy texture features and
their link to AGC had limited to no success. Our analysis provides
insights into the rate at which deforestation and reforestation has
taken place over a multi-decadal scale in the central Congo basin. As
such, it provides a useful historical context while interpreting past
and ongoing forest research in the area.

>>>>

\textbf{Tropical ecosystem services are severely impacted by
deforestation and forest degradation
\citep{hansen2013, houghton2012, tyukavina2015}. Not only does tropical
forest Land-Use and Land-Cover Change (LULCC) constitute 10 to 15\% of
the total global carbon emissions \citep{vanderwerf2009}, changes in
forest fragmentation affect the forest structure and function
\citep{fauset2017, brinck2017, didham1999}. Strong fragmentation effects
decreases the number of large trees along forest edges
\citep{laurance2000, magnago2015}, while species composition and
biodiversity are equally negatively affected
\citep{poorter2006, barlow2016, vandeperre2018}. Estimates show that
31\% of carbon emissions are caused by edge effects alone
\citep{brinck2017}.}

\textbf{Accurate estimates of LULCC and forest canopy structure are
therefore imperative to estimate carbon emissions and other ecosystem
services \citep{houghton2012, hansen2013}. Remote sensing products have
been key inputs in LULCC assessments as they provide accurate spatial
information to help estimate carbon emissions
\citep{mitchard2018, hansen2013}. More so, high resolution aerial images
provide scientists tools to monitor forest extent, structure and carbon
emissions as canopy texture is linked to above ground biomass
\citep{Ploton2012, Couteron2005, Barbier2010}. Yet, most of these
estimates are limited in time to recent decades
\citep{defries2002, houghton2012, hansen2013, achard2014}.}

\textbf{Historical estimates of Land-Use and Land-Cover (LULC), in the
pre-satellite era (\textless{} 1972), exist but generally rely on
non-spatially explicit data (i.e socio-economic data)
\citep{houghton2012, ramankutty1999, defries2002, sader1988}. Efforts
have been made to use other geo-spatial data sources such as historical
maps \citep{willcock2016}, declassified CORONA satellite surveillance
data across the US and central Brazil \citep{song2015} as well as aerial
surveys in post World War II German \citep{nita2018}. Survey data across
the African continent is less common, inaccessible or both. Some studies
do exist, as \citet{buitenwerf2012} and \citet{hudak1998} used aerial
survey images to map vegetation changes in South African savannas,
whilst \citet{frankl2015} and \citet{nyssen2016} mapped the Ethiopian
highlands of 1930's.}

\textbf{Across the central Congo Basin most of these historical images
were collected within the context of national cartographic efforts by
the ``Institut Géographique du Congo Belge'' in Kinshasa (then
Léopoldville), DR Congo. Despite the existence of large archives of
aerial survey imagery (Figure 1, Appendix Figure 3), as of yet, no
studies have valorized these data. The lack of a consistent valorization
effort is unfortunate as the African rainforest is the second largest on
Earth and covers \textasciitilde{}630 million ha, representing up to 66
Pg of carbon storage \citep{Lewis2009b}, and currently loses forest at
an increasing pace \citep{butsic2015}. Given the impact of LULCC on the
structure and functioning of central African tropical forests, and their
influence on both carbon dynamics \citep{Kearsley2013} and biodiversity
\citep{vandeperre2018}, accurate long term reporting of historical
forest cover warrants more attention \citep{willcock2016}.}

\textbf{Here, we use a combination of historical aerial photography and
contemporary remote sensing data to map long-term changes in the extent
and structure of the tropical forest surrounding Yangambi (DR Congo), in
the central Congo Basin, effectively linking the start of the
antrophocene \citep{lewis2015} with current assessments. Yangambi was,
and remains, a focal center of forest and agricultural research and
development in the central Congo Basin. Past research in the region
allows for thorough assessment of land-use and land-cover change from a
multi-disciplinary point of view, confronting us with complex
deforestation and land-use patterns.}

\textbf{We leverage Structure-from-Motion to generate a large
orthomosaic of historical imagery and develop a Convolutional Neural
Network based forest cover mapping approach, based upon a
semi-supervised generated dataset extensively leveraging data
augmentation. Our methodology aims to provides a historical insight into
important LULCC spatial patterns in Yangambi, such as fragmentation and
edge complexity. We further contextualize the influence of changes in
the forest's life history on past and current research into Above Ground
Carbon (AGC) storage \citep{Kearsley2013} and biodiversity
\citep{vandeperre2018} in the central Congo Basin. Our fast scalable
mapping approach for historical aerial survey data, using limited
supervised input, would further support long-term land-use and
land-cover change analysis across the central Congo Basin.}

@@

We automatically delineated all natural forest in the historical data,
thus excluding tree plantations, thinned or deteriorated forest stands
which \textbf{showed visible canopy cover loss}, fields and buildings.
We used the Unet Convolutional Neural Net (CNN, \citet{ronneberger2015})
architecture implemented in Keras \citep{chollet2015} with an
efficientnetb3 pre-processing backbone \citep{yakubovskiy2019} running
on TensorFlow \citep{martinabadi2015} to train a binary classifier
(i.e.~forest or non-forested). Training data were collected from the
orthomosaic by randomly selecting 513 pixel square tiles from
\textbf{locations within} homogeneous forested or non-forested polygons
in the historical orthomosaic \textbf{(Figure 5). Separate polygons were
selected for training, testing and validation purposes. Validation
polygons were sampled 300 times, while both testing and validation
polygons were sampled at 100 random locations. Tiles extracted from
locations close to the polygon border at times contained mixed cover
types. Tiles with mixed cover types were removed from the list of source
tiles (Table 1).} Homogeneous \textbf{source} tiles were combined in
synthetic landscapes using a random gaussian field based binary mask
(Figure 3). We generated 5000 synthetic landscapes \textbf{(balancing
forest and non-forest classes)} for training, while 500 landscapes were
generated for both the validation and the testing datasets for a total
of 6000 synthetic landscapes. In order to limit stitch line
misclassifications, along the seams of mosaicked images, we created
synthetic landscapes with different forest tiles to mimick forest
texture transitions. We applied this technique to 10\% of the generated
synthetic landscapes \textbf{(across training, validation and testing
data)}.

@@

To quantify changes in the structure of forest cover and its
disturbances we used spatial pattern analysis landscape (fragmentation)
metrics \citep{dale1999}. \textbf{Landscape metrics provide a
mathematical framework for the analysis of discrete land‐cover classes
and allows us to capture their composition and configuration. These
metrics are therefore commonly used to compare how landscapes change
over time \citep{hesselbarth2019}. In particular, fractals provide a way
to quantify complex natural landscapes, including their self-similarity,
across scales \citep{Li2000, Mandelbrot1977}} We report the ratio of
edge to area and the fractal dimension to quantify landscape complexity
of forest disturbances. \textbf{A fractal dimension closer to 2 suggest
a more complex (fragmented) landscape.}

Statistics were calculated for all forest disturbance patches larger
than 1 ha and smaller than the 95\textsuperscript{th} percentile of the
patch size distribution using the R package landscapemetrics
\citep{hesselbarth2019}. We provide mean and standard deviation on edge,
area, their ratio and fractal dimension for both the historical and
contemporary \citet{hansen2013} forest cover maps.

@@

Over the past half century there has been a clear shift in land use in
Yangambi (Figure 5). Land use has shifted away from for example a
regular fishbone deforestation pattern that emerges when (large scale)
agricultural interests dominate the landscape \citep{arima2015}, to a
more fragmented landscape (Figure 5D). The former is consistent with
historical land management at the time of the aerial survey
\citep{bustillo2018}. These regular patterns have since been reversed
because of a decrease in large scale intensive agriculture, replaced by
ad-hoc small scale subsistence farming with large perimeter to area
relationships (i.e.~ragged edges). Consequently, edge effects in the
current landscape are far more pronounced than in the historical scene.

@@

\textbf{Given the impact of tropical forest disturbances on atmospheric
carbon emissions, biodiversity and ecosystem productivity accurate long
term reporting of LULCC is an imperative. Our analysis of historical
aerial survey images (1958) of the Central Congo Basin provides a window
into the state of the forest at the start of the anthropocene. The use
of a CNN based LULC classifier, using synthetic landscapes based image
augmentation, provides a robust semi-supervised solution which scales
across space and time, even for image scenes with inconsistent
illumination or sharpness. Combined with contemporary remote sensing
data we have shown that historical aerial survey data can be used to
quantify long-term changes in LULC and AGC. We showed a shift from
previously highly structured industrial deforestation of large areas for
plantation purposes, to discrete smallholder clearing for farming,
increasing landscape fragmentation but also opportunties for substantial
regrowth. Efforts to quantify canopy texture features and their link to
AGC had limited to no success. Our analysis provides insights into the
rate at which deforestation and reforestation has taken place over a
multi-decadal scale in the central Congo basin. As such, it provides a
useful historical context while interpreting past and ongoing forest
research in the area.}
